Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2023-10-05 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Update case3

Table 1: Concepts in multilevel methods

| **Concept**         | **Details**                                                                   |
|---------------------|-------------------------------------------------------------------------------|
| Sample unit         | Units of the population from which measurements are taken in a sample.        |
| Hierarchical levels | Data are grouped into clusters at different levels. A three-level             |
| Fixed effect        | Here we assume that the values of an independent variable are fixed, i.e.,    |
|                     | the values observed in the study are representative of all values in the      |
|                     | in the dependent variable y e.g., blood pressure between treatments A and B.  |
| Random effect       | The values of an independent variable are assumed to be randomly drawn from   |
|                     | admission we might select only certain hospitals that are representative of   |
|                     | the difference of y between individual hospitals, but rather the variation of |
| ICC                 | The variability due to clustering is often measured by means of the           |
|                     | intraclass coefficient (ICC). The ICC can be seen as the percentage           |
|                     | of variance that can be attributed to the cluster-level, where a high         |
|                     | ICC would indicate that a lot of variability is due to the cluster            |
|                     | structure.                                                                    |
| Random effect       | Multilevel models typically accommodate for variability by including          |
|                     | a separate group mean for each cluster. In addition to random                 |
|                     | intercepts, multilevel models can also include random coefficients            |
|                     | and heterogeneous residual error variances across clusters [see e.g.          |
|                     | @gelm06, @hox17 and @jong21]. [TODO: add stratified intercept as concept..]                     |

@@

- `popmis` from the \pkg{mice} package [simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools with data that are MAR, @mice];
- `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies with data that are MAR, @metamisc];
- `obesity` from the \pkg{miceheckman} package [simulated data on obesity, $n = 2,111$ patients across $N = 5$ regions with data that are MNAR].

@@

Install non-CRAN packages if necessary:
```{r packages, eval=FALSE, include=TRUE}
devtools::install_github("amices/ggmice")
# devtools::install_github("hanneoberman/miceheckman")
```

@@

TODO: explain exclusion restriction.

Data are simulated and included in the `miceheckman` package. We will use the following variables:

- `region` Cluster variable,
- `gender` Gender (0=male, 1=female),
- `age` Age of the patient,
- `height` Height in meters,
- `weight` Weight in kilograms,
- `time`` Response time in minutes (inclusion-restriction variable).

Load the data:
```{r obesity}
data("data_heckman", package = "micemd")
```

Select relevant variables:
```{r bmi}
dat <- data_heckman
```

Visualize missing data pattern:
```{r obesity-md}
plot_pattern(dat)
```

@@

>>>>

```{r clus,  echo = FALSE, warning = FALSE,cache = TRUE}
concepts <- data.frame(Concept=c("Sample unit","Hierarchical levels","Fixed effect","Random effect","Mixed effect","ICC","Stratified intercept"),
                       Details=c("Units of the population from which measurements are taken in a sample.",
                                 "Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.",
                                 "Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as the administration of a drug.",
                                 "Effects that are a source of random variation in the data, and whose levels are not fully sampled. e.g. individuals are drawn from a population of hospitals, here it is not possible to sample all hospitals but drug effects could vary between hospitals.",
                                 "Includes fixed and random effects, e.g. the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated. Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals. In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g. @gelm06, @hox17 and @jong21).",
                                 "The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.",
                                 ""))
library(kableExtra)
concepts%>%
  kable( booktabs = T, escape = F,
         caption = "Concepts in multilevel methods",
         col.names = c("Concept","Details"), align = "l")%>%
  column_spec(c(1), width = "3cm")%>%
  column_spec(c(2), width = "12cm")
```

@@

- `popmis` from the \pkg{mice} package [simulated data on perceived popularity, $n = 2,000$ pupils across $N = 100$ schools with data that are MAR, @mice];
- `impact` from the \pkg{metamisc} package [empirical data on traumatic brain injuries, $n = 11,022$ patients across $N = 15$ studies with data that are MAR, @metamisc];
- `obesity` from the \pkg{micemd} package [simulated data on obesity, $n = 2,111$ patients across $N = 5$ regions with data that are MNAR].

@@

Install non-CRAN packages if necessary:
```{r packages, eval=FALSE, include=TRUE}
devtools::install_github("amices/ggmice")

```

@@

In this example, we demonstrate a multilevel imputation of an intercept and slope random model with a continuous response. We utilize the obesity dataset included in the `micemd` package, a simulated dataset that emulates an electronic survey in which individuals are asked to provide information about their weight and consumption habits in different countries. To easy the explanations we simulate data for 5 clusters and we reduce the dataset to the following variables:

- `region` Cluster variable,
- `gender` Gender (0=male, 1=female),
- `age` Age of the patient,
- `height` Height in meters,
- `weight` Weight in kilograms,
- `time`` Response time in minutes (inclusion-restriction variable).
- `FamOb`` Family obesity history (yes or not)

Age and FamOb are missing according to a Missing At Random (MAR) mechanism, whereas the weight variable is subject to selection bias, following an indirect Missing Not At Random (MNAR) mechanism, that usually happens when it is unobserved or omitted a variable affecting both the value of incomplete variable (here weight) and its probability of missing.

The analysis model considers BMI as the dependent variable, predicted by Age, Height, Gender, and FAV. Random slopes are considered for the Age variable, which contains missing values.

The model is represented as:
$$BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FAV_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}$$
We load the data:
```{r obesity}
data("data_heckman", package = "micemd")
dat <- data_heckman
```

Now, let's begin by examining the missing patterns in the data by cluster:
```{r obesity-md}
plot_pattern(dat)
```

We observe that the proportion of missingness for the Age and FAVC variables is quite similar across the clusters. However, regarding the weight variable, we notice a heterogeneous proportion of missingness, with systematic missingness in cluster 5.
In order to evaluate if it is required a imputation method for 1-level or 2-level we assess the Intraclass Correlation Coefficient (ICC) for the outcome variable, we use the "performance" package:

```{r obesity icc}
dat$bmi=dat$Weight/(dat$Height)^2
Nulmodel <- lme4::lmer(bmi ~ 1 + (1|Cluster), data = dat)
#Performance::ICC(Nulmodel)
```

Since the ICC is above 0.1, we decide to employ two-level (2l) imputation methods. In this imputation process, we include all predictor variables in the main model and since BMI is a composite variable. We also incorporate weight and height variables.

To determine the best imputation method for these variables, we can use the `find.defaultMethod`function provided in the `micemd` package, which suggests an appropriate method for MAR variables based on the type of variable, number of observations in the cluster, and number of clusters. It suggests using "2l.2stage.bin" for the FAV variable and "2l.2stage.norm" for the age and weight variables. However, after inspecting the age density plot, we consider modifying its method to "2l.2stage.ppm." For the BMI variable, we employ deterministic imputation.
```{r obesity meth}
meth_mar <- find.defaultMethod(dat, ind.clust=7, I.small = 7, ni.small = 100, prop.small = 0.4)
```
For these imputation models, it is necessary to specify the prediction matrix, with the cluster variable labeled as -2 and the predictor variable labeled as 2, encompassing all variables.

After confirming convergence, we proceed to fit the main model and save the results for future use. Users are considering the possibility that the weight variable may not have been selected randomly. It's likely that an omitted variable, like self-esteem, could influence this selection. For instance, individuals with lower self-esteem might have higher weight values, impacting their willingness to provide honest information due to embarrassment.

To address this situation, two approaches have been proposed for dealing with Missing Not at Random (MNAR) data: pattern-mixed models and selection models. Within pattern-mixed models, methods like the delta method and more advanced ones like NARFS have been suggested. In contrast, the selection model approach includes methods such as the Heckman model, which can be particularly useful in this case. Several methods, including Galimar2017, Hammon2021, and the recent Munoz method, designed for two-level data, allowing for variations in intercepts and exposure effects (random intercept and slope).

@@

To apply the Munoz method, the weight variable should be specified as "2l.2stage.heckman" found in the micemd package. Additionally, the prediction matrix needs modification because this method involves specifying two equations: one for the outcome, describing the incomplete variable in terms of partially observed predictors (in this case, all variables from the main model), and the other for the selection model, explaining the probability of being observed based on certain variables.

TODO: explain exclusion restriction.
In this specific case, it's assumed that height does not influence selection (-4), and an exclusion restriction variable is introduced as an instrumental variable affecting selection but not the outcome. Time of response is used for this purpose, as it's believed that subjects taking more time to complete the survey are more likely to not respond due to poor connectivity or lack of familiarity with electronic surveys (specified as -3).

After implementing these imputation procedures, it's crucial to check for convergence. If convergence issues arise and imputed values are far from plausible, especially affecting BMI values, a post-imputation procedure is employed to constrain BMI values within a plausible range.

Then after this modification we proceed to compare the effects on the model
###Plot estimates

Visualize missing data pattern:

### 2023-10-05 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Update case3

\Abstract{
This tutorial illustrates the imputation of incomplete multilevel data
with the \proglang{R} packackage \pkg{mice}.
}

@@

Table 1: Concepts in multilevel methods

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.2100}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 2\tabcolsep) * \real{0.7900}}@{}}
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
\textbf{Concept}
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\textbf{Details}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Sample unit & Units of the population from which measurements are taken
in a sample. \\
Hierarchical levels & Data are grouped into clusters at different
levels. A three-level \\
Fixed effect & Here we assume that the values of an independent variable
are fixed, i.e., \\
& the values observed in the study are representative of all values in
the \\
& in the dependent variable y e.g., blood pressure between treatments A
and B. \\
Random effect & The values of an independent variable are assumed to be
randomly drawn from \\
& admission we might select only certain hospitals that are
representative of \\
& the difference of y between individual hospitals, but rather the
variation of \\
ICC & The variability due to clustering is often measured by means of
the \\
& intraclass coefficient (ICC). The ICC can be seen as the percentage \\
& of variance that can be attributed to the cluster-level, where a
high \\
& ICC would indicate that a lot of variability is due to the cluster \\
& structure. \\
Random effect & Multilevel models typically accommodate for variability
by including \\
& a separate group mean for each cluster. In addition to random \\
& intercepts, multilevel models can also include random coefficients \\
& and heterogeneous residual error variances across clusters {[}see
e.g. \\
& \citet{gelm06}, \citet{hox17} and \citet{jong21}{]}. {[}TODO: add
stratified intercept as concept..{]} \\
\end{longtable}

@@

\begin{itemize}
\tightlist
\item
  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
  perceived popularity, \(n = 2,000\) pupils across \(N = 100\) schools
  with data that are MAR,][]{mice};
\item
  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
  studies with data that are MAR,][]{metamisc};
\item
  \texttt{obesity} from the \pkg{miceheckman} package {[}simulated data
  on obesity, \(n = 2,111\) patients across \(N = 5\) regions with data
  that are MNAR{]}.
\end{itemize}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> devtools::install_github("amices/ggmice")
R> # devtools::install_github("hanneoberman/miceheckman")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
\end{CodeInput}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect  group term  estimate std.error statistic   p.value conf.low conf.high
   <chr>   <chr> <chr>    <dbl>     <dbl>     <dbl>     <dbl>    <dbl>     <dbl>
 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   2.99e-31   0.0571     0.130
 2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
 3 fixed   <NA>  age     1.03     0.00265     12.9   7.40e-38   1.03       1.04 
 4 fixed   <NA>  moto~   0.651    0.0732      -3.82  1.34e- 4   0.522      0.811
 5 fixed   <NA>  moto~   0.489    0.0555      -6.30  2.97e-10   0.391      0.611
 6 fixed   <NA>  moto~   0.274    0.0321     -11.0   2.28e-28   0.218      0.345
 7 fixed   <NA>  pupi~   3.20     0.317       11.7   8.18e-32   2.63       3.88 
 8 fixed   <NA>  pupi~   1.75     0.195        5.06  4.27e- 7   1.41       2.18 
 9 fixed   <NA>  ctIII   2.41     0.268        7.89  3.05e-15   1.94       2.99 
10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.55e-19   1.92       2.76 
11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- imp %>% 
+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
R> tidy(pool(fit))
\end{CodeInput}
\begin{CodeOutput}
                     term    estimate   std.error  statistic      p.value
1             (Intercept) -2.35199614 0.340174191  -6.914093 4.993068e-12
2                    type -0.41267936 0.180271952  -2.289204 2.208652e-02
3                     age  0.03049024 0.001570154  19.418637 1.233010e-81
4 as.factor(motor_score)2 -0.66765325 0.068737838  -9.713038 3.478858e-22
5 as.factor(motor_score)3 -1.05520488 0.070218678 -15.027410 2.536132e-50
6 as.factor(motor_score)4 -1.51238928 0.072304221 -20.917026 1.857373e-90
7                   pupil  0.48421506 0.038982924  12.421209 6.781796e-17
8                      ct  0.43474556 0.029968441  14.506779 2.341875e-36
             b          df dfcom         fmi      lambda m         riv
1 5.301620e-04 10113.60453 11013 0.005694385 0.005497777 5 0.005528170
2 4.905995e-05 10892.82806 11013 0.001994781 0.001811557 5 0.001814845
3 3.333628e-08  6323.59900 11013 0.016537098 0.016226102 5 0.016493731
4 4.194026e-05  8322.17049 11013 0.010889419 0.010651742 5 0.010766424
5 5.136864e-05  7631.13821 11013 0.012760548 0.012501842 5 0.012660117
6 1.403437e-04  2830.62134 11013 0.032897234 0.032214161 5 0.033286456
7 3.571667e-04    49.96887 11013 0.309144148 0.282035215 5 0.392825972
8 8.586575e-05   294.70102 11013 0.120676307 0.114728924 5 0.129597506
          ubar
1 1.150823e-01
2 3.243910e-02
3 2.425379e-06
4 4.674562e-03
5 4.869020e-03
6 5.059488e-03
7 1.091068e-03
8 7.950686e-04
\end{CodeOutput}
\begin{CodeInput}
R> as.mitml.result(fit)
\end{CodeInput}
\begin{CodeOutput}
[[1]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10495.423 10561.192 -5238.712 10477.423     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2843  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.37194                 -0.41015                  0.03052  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65802                 -1.04611                 -1.51245  
                  pupil                       ct  
                0.50405                  0.42496  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

@@

TODO: explain exclusion restriction.

Data are simulated and included in the \texttt{miceheckman} package. We
will use the following variables:

\begin{itemize}
\tightlist
\item
  \texttt{region} Cluster variable,
\item
  \texttt{gender} Gender (0=male, 1=female),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{height} Height in meters,
\item
  \texttt{weight} Weight in kilograms,
\item
  `time`` Response time in minutes (inclusion-restriction variable).
\end{itemize}

Load the data:

\begin{CodeChunk}
\begin{CodeInput}
R> data("data_heckman", package = "micemd")
\end{CodeInput}
\end{CodeChunk}

Select relevant variables:

\begin{CodeChunk}
\begin{CodeInput}
R> dat <- data_heckman
\end{CodeInput}
\end{CodeChunk}

@@

@@

>>>>

\Abstract{
This tutorial illustrates the imputation of incomplete multilevel data
with the \proglang{R} packackage \pkg{mice}. Our scope is only simple
multilevel models, to show how imputation can yield less biased
estimates from incomplete clustered data. More complex models can be
accomodated, but are outside the scope of this paper.
}

@@

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:clus}Concepts in multilevel methods}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{3cm}>{\raggedright\arraybackslash}p{12cm}}
\toprule
Concept & Details\\
\midrule
Sample unit & Units of the population from which measurements are taken in a sample.\\
Hierarchical levels & Data are grouped into clusters at different levels, observations belonging to the same cluster are expected to share certain characteristics.\\
Fixed effect & Effects that are constant across all sample units, e.g. something that researchers control for and can repeat, such as the administration of a drug.\\
Random effect & Effects that are a source of random variation in the data, and whose levels are not fully sampled. e.g. individuals are drawn from a population of hospitals, here it is not possible to sample all hospitals but drug effects could vary between hospitals.\\
Mixed effect & Includes fixed and random effects, e.g. the fixed effect would be the treatment effect of a drug and the random effect would be the ID of the hospital where the patient is treated. Multilevel models typically accommodate for variability by including a separate group mean for each cluster e.g random intercept on hospitals. In addition to random intercepts, multilevel models can also include random coefficients and heterogeneous residual error variances across clusters (see e.g. @gelm06, @hox17 and @jong21).\\
\addlinespace
ICC & The variability due to clustering is often measured by means of the intraclass coefficient (ICC). The ICC can be seen as the percentage
of variance that can be attributed to the cluster-level, where a high ICC would indicate that a lot of variability is due to the cluster structure.\\
Stratified intercept & \\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{itemize}
\tightlist
\item
  \texttt{popmis} from the \pkg{mice} package \citep[simulated data on
  perceived popularity, \(n = 2,000\) pupils across \(N = 100\) schools
  with data that are MAR,][]{mice};
\item
  \texttt{impact} from the \pkg{metamisc} package \citep[empirical data
  on traumatic brain injuries, \(n = 11,022\) patients across \(N = 15\)
  studies with data that are MAR,][]{metamisc};
\item
  \texttt{obesity} from the \pkg{micemd} package {[}simulated data on
  obesity, \(n = 2,111\) patients across \(N = 5\) regions with data
  that are MNAR{]}.
\end{itemize}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> devtools::install_github("amices/ggmice")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
\end{CodeInput}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect  group term  estimate std.error statistic   p.value conf.low conf.high
   <chr>   <chr> <chr>    <dbl>     <dbl>     <dbl>     <dbl>    <dbl>     <dbl>
 1 fixed   <NA>  (Int~   0.0863   0.0182     -11.6   3.00e-31   0.0571     0.130
 2 fixed   <NA>  type~   0.757    0.137       -1.54  1.22e- 1   0.531      1.08 
 3 fixed   <NA>  age     1.03     0.00265     12.9   7.40e-38   1.03       1.04 
 4 fixed   <NA>  moto~   0.651    0.0732      -3.82  1.34e- 4   0.522      0.811
 5 fixed   <NA>  moto~   0.489    0.0555      -6.30  2.97e-10   0.391      0.611
 6 fixed   <NA>  moto~   0.274    0.0321     -11.0   2.28e-28   0.218      0.345
 7 fixed   <NA>  pupi~   3.20     0.317       11.7   8.18e-32   2.63       3.88 
 8 fixed   <NA>  pupi~   1.75     0.195        5.06  4.27e- 7   1.41       2.18 
 9 fixed   <NA>  ctIII   2.41     0.268        7.89  3.05e-15   1.94       2.99 
10 fixed   <NA>  ctIV~   2.30     0.214        8.95  3.56e-19   1.92       2.76 
11 ran_pa~ name  sd__~   0.230   NA           NA    NA         NA         NA    
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- imp %>% 
+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
R> tidy(pool(fit))
\end{CodeInput}
\begin{CodeOutput}
                     term    estimate   std.error  statistic      p.value
1             (Intercept) -2.35203726 0.340181747  -6.914061 4.994037e-12
2                    type -0.41265892 0.180274846  -2.289054 2.209524e-02
3                     age  0.03049023 0.001570162  19.418521 1.238416e-81
4 as.factor(motor_score)2 -0.66764920 0.068737865  -9.712975 3.480413e-22
5 as.factor(motor_score)3 -1.05520001 0.070218940 -15.027285 2.540225e-50
6 as.factor(motor_score)4 -1.51238349 0.072304262 -20.916934 1.850073e-90
7                   pupil  0.48421447 0.038982800  12.421234 6.772069e-17
8                      ct  0.43474621 0.029968474  14.506785 2.342253e-36
             b          df dfcom         fmi      lambda m         riv
1 5.281599e-04 10119.70041 11013 0.005673265 0.005476771 5 0.005506932
2 4.881699e-05 10893.89378 11013 0.001985736 0.001802528 5 0.001805783
3 3.335358e-08  6320.89582 11013 0.016545467 0.016234340 5 0.016502243
4 4.188703e-05  8327.24771 11013 0.010875748 0.010638213 5 0.010752602
5 5.135721e-05  7632.20045 11013 0.012757637 0.012498967 5 0.012657168
6 1.403058e-04  2831.75462 11013 0.032888241 0.032205434 5 0.033277139
7 3.571497e-04    49.97295 11013 0.309130930 0.282023647 5 0.392803531
8 8.586645e-05   294.69765 11013 0.120677044 0.114729598 5 0.129598366
          ubar
1 1.150898e-01
2 3.244044e-02
3 2.425385e-06
4 4.674630e-03
5 4.869071e-03
6 5.059539e-03
7 1.091079e-03
8 7.950697e-04
\end{CodeOutput}
\begin{CodeInput}
R> as.mitml.result(fit)
\end{CodeInput}
\begin{CodeOutput}
[[1]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10495.423 10561.192 -5238.712 10477.423     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2843  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.37195                 -0.41014                  0.03052  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65802                 -1.04611                 -1.51245  
                  pupil                       ct  
                0.50405                  0.42496  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

@@

In this example, we demonstrate a multilevel imputation of an intercept
and slope random model with a continuous response. We utilize the
obesity dataset included in the \texttt{micemd} package, a simulated
dataset that emulates an electronic survey in which individuals are
asked to provide information about their weight and consumption habits
in different countries. To easy the explanations we simulate data for 5
clusters and we reduce the dataset to the following variables:

\begin{itemize}
\tightlist
\item
  \texttt{region} Cluster variable,
\item
  \texttt{gender} Gender (0=male, 1=female),
\item
  \texttt{age} Age of the patient,
\item
  \texttt{height} Height in meters,
\item
  \texttt{weight} Weight in kilograms,
\item
  `time`` Response time in minutes (inclusion-restriction variable).
\item
  `FamOb`` Family obesity history (yes or not)
\end{itemize}

Age and FamOb are missing according to a Missing At Random (MAR)
mechanism, whereas the weight variable is subject to selection bias,
following an indirect Missing Not At Random (MNAR) mechanism, that
usually happens when it is unobserved or omitted a variable affecting
both the value of incomplete variable (here weight) and its probability
of missing.

The analysis model considers BMI as the dependent variable, predicted by
Age, Height, Gender, and FAV. Random slopes are considered for the Age
variable, which contains missing values.

The model is represented as:
\[BMI_{ij}= (\beta_{o}+ b_{oj} ) + (\beta_{1}+ b_{oj})* Age_{ij} + \beta_{2}*FAV_{ij}+ \beta_{3}Gender_{ij} + \epsilon_{ij}\]
We load the data:

\begin{CodeChunk}
\begin{CodeInput}
R> data("data_heckman", package = "micemd")
R> dat <- data_heckman
\end{CodeInput}
\end{CodeChunk}

Now, let's begin by examining the missing patterns in the data by
cluster:

@@

We observe that the proportion of missingness for the Age and FAVC
variables is quite similar across the clusters. However, regarding the
weight variable, we notice a heterogeneous proportion of missingness,
with systematic missingness in cluster 5. In order to evaluate if it is
required a imputation method for 1-level or 2-level we assess the
Intraclass Correlation Coefficient (ICC) for the outcome variable, we
use the ``performance'' package:

\begin{CodeChunk}
\begin{CodeInput}
R> dat$bmi=dat$Weight/(dat$Height)^2
R> Nulmodel <- lme4::lmer(bmi ~ 1 + (1|Cluster), data = dat)
R> #Performance::ICC(Nulmodel)
\end{CodeInput}
\end{CodeChunk}

Since the ICC is above 0.1, we decide to employ two-level (2l)
imputation methods. In this imputation process, we include all predictor
variables in the main model and since BMI is a composite variable. We
also incorporate weight and height variables.

To determine the best imputation method for these variables, we can use
the \texttt{find.defaultMethod}function provided in the \texttt{micemd}
package, which suggests an appropriate method for MAR variables based on
the type of variable, number of observations in the cluster, and number
of clusters. It suggests using ``2l.2stage.bin'' for the FAV variable
and ``2l.2stage.norm'' for the age and weight variables. However, after
inspecting the age density plot, we consider modifying its method to
``2l.2stage.ppm.'' For the BMI variable, we employ deterministic
imputation.

\begin{CodeChunk}
\begin{CodeInput}
R> meth_mar <- find.defaultMethod(dat, ind.clust=7, I.small = 7, ni.small = 100, prop.small = 0.4)
\end{CodeInput}
\end{CodeChunk}

For these imputation models, it is necessary to specify the prediction
matrix, with the cluster variable labeled as -2 and the predictor
variable labeled as 2, encompassing all variables.

After confirming convergence, we proceed to fit the main model and save
the results for future use. Users are considering the possibility that
the weight variable may not have been selected randomly. It's likely
that an omitted variable, like self-esteem, could influence this
selection. For instance, individuals with lower self-esteem might have
higher weight values, impacting their willingness to provide honest
information due to embarrassment.

To address this situation, two approaches have been proposed for dealing
with Missing Not at Random (MNAR) data: pattern-mixed models and
selection models. Within pattern-mixed models, methods like the delta
method and more advanced ones like NARFS have been suggested. In
contrast, the selection model approach includes methods such as the
Heckman model, which can be particularly useful in this case. Several
methods, including Galimar2017, Hammon2021, and the recent Munoz method,
designed for two-level data, allowing for variations in intercepts and
exposure effects (random intercept and slope).

@@

To apply the Munoz method, the weight variable should be specified as
``2l.2stage.heckman'' found in the micemd package. Additionally, the
prediction matrix needs modification because this method involves
specifying two equations: one for the outcome, describing the incomplete
variable in terms of partially observed predictors (in this case, all
variables from the main model), and the other for the selection model,
explaining the probability of being observed based on certain variables.

TODO: explain exclusion restriction. In this specific case, it's assumed
that height does not influence selection (-4), and an exclusion
restriction variable is introduced as an instrumental variable affecting
selection but not the outcome. Time of response is used for this
purpose, as it's believed that subjects taking more time to complete the
survey are more likely to not respond due to poor connectivity or lack
of familiarity with electronic surveys (specified as -3).

After implementing these imputation procedures, it's crucial to check
for convergence. If convergence issues arise and imputed values are far
from plausible, especially affecting BMI values, a post-imputation
procedure is employed to constrain BMI values within a plausible range.

Then after this modification we proceed to compare the effects on the
model \#\#\#Plot estimates
